---
title: "Pratical Machine Learining Project"
author: "Fang Tan"
date: "Sunday, October 9, 2016"
output:
  html_document:
    theme: spacelab
---
Quantified Self devices not only can quantify how much of a particular activity one does, but also quantify how well they do it. Due to the relatively inexpensive data collection, it is highly valued by a lot of companies to help improve health.

The focus of this project is to utilize sampled data to build a model to predict the manner in which they did the exercise.


### Libraries
The following libraries were used throughout the code.
```{r}
library(ggplot2)
library(lattice)
library(caret)
library(corrplot)
library(knitr)
library(kernlab)
library(randomForest)
```


### Loading and preprocessing the data
The data for this project come from this Human Activity Recognition (HAR): [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har). Thank them for sharing their data.

The data set were downloaded from:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

```{r, eval = FALSE}
# file URL and destination file
fileUrl1 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
destfile1 <- "./data/pml-training.csv"
fileUrl2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
destfile2 <- "./data/pml-testing.csv"

# download the file
download.file(fileUrl1, destfile = destfile1)
download.file(fileUrl2, destfile = destfile2)
```

The training data was then loaded into R.

```{r}
# read the csv file for training 
data_training <- read.csv("./pml-training.csv", na.strings= c("NA",""," "))
```

There was a lot of NA values in the data which would create a lot of noise for the model. As a result, these columns were removed from the data set. The first eight columns that acted as identifiers for the experiment were also removed.

```{r}
# clean the data by removing columns with NAs etc
data_training_NAs <- apply(data_training, 2, function(x) {sum(is.na(x))})
data_training_clean <- data_training[,which(data_training_NAs == 0)]

# remove identifier columns such as name, timestamps etc
data_training_clean <- data_training_clean[8:length(data_training_clean)]
```

##Splitting the training sampling for Cross-Validation

Although there is a unknown testing data set, however, the unkown testing data set has a small sampling size which is 20. We want to split the large training samples to a training set and validation set as suggested in Week 1 class.


Since we have a large amount of observations, 60% is used for the training set and 40% is used for the validation set. The data is randomly split into these two parts. Seed was set for reproduction purpose.

```{r}
# split the cleaned testing data into training and cross validation
inTrain <- createDataPartition(y = data_training_clean$classe, p = 0.6, list = FALSE)
training <- data_training_clean[inTrain, ]
crossval <- data_training_clean[-inTrain, ]
```

##Choose the model
A random forest model was selected to predict the classification because it has methods for balancing error in class population unbalanced data sets. The correlation between any two trees in the forest increases the forest error rate. Therefore, a correllation plot was produced in order to see how strong the variables relationships are with each other.

```{r, fig.height = 5, fig.width = 7}
# plot a correlation matrix
correlMatrix <- cor(training[, -length(training)])
corrplot(correlMatrix, order = "FPC", method = "circle", type = "lower", tl.cex = 0.8,  tl.col = rgb(0, 0, 0))
```

In this type of plot the dark red and blue colours indicate a highly negative and positive relationship respectively between the variables. There isn't much concern for highly correlated predictors which means that all of them can be included in the model.

Then a model was fitted with the outcome set to the training class and all the other variables used to predict.

```{r}
# fit a model to predict the classe using everything else as a predictor
model <- randomForest(classe ~ ., data = training)
model
```

The model produced a very small OOB error rate of .56%. This was deemed satisfactory enough to progress the testing.

## Cross-validation
The model was then used to classify the remaining 30% of data. The results were placed in a confusion matrix along with the actual classifications in order to determine the accuracy of the model.

```{r}
# crossvalidate the model using the remaining 30% of data
predictCrossVal <- predict(model, crossval)
confusionMatrix(crossval$classe, predictCrossVal)
```

This model yielded a 99.3% prediction accuracy. Again, this model proved very robust and adequete to predict new data.

## Predictions
A separate data set was then loaded into R and cleaned in the same manner as before. The model was then used to predict the classifications of the 20 results of this new data.

```{r}
# apply the same treatment to the final testing data
data_test <- read.csv("./pml-testing.csv", na.strings= c("NA",""," "))
data_test_NAs <- apply(data_test, 2, function(x) {sum(is.na(x))})
data_test_clean <- data_test[,which(data_test_NAs == 0)]
data_test_clean <- data_test_clean[8:length(data_test_clean)]

# predict the classes of the test set
predictTest <- predict(model, data_test_clean)
predictTest
```

## Conclusions
With the abundance of information given from multiple measuring instruments it's possible to accurately predict how well a person is preforming an excercise using a relatively simple model. 